Challenges to the Two-Infall Scenario by Large Stellar Age Catalogs

### 2025-05-23 src/tex/ms.tex
Manuscript updates

% Author list
% \author[]{Amanda Ash}
\author[0000-0003-3781-0747]{Liam O.\ Dubay}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
\author[0000-0001-7258-1834]{Jennifer A.\ Johnson}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
\author[0000-0002-6534-8783]{James W.\ Johnson}
\affiliation{Observatories of the Carnegie Institution for Science, 813 Santa Barbara St., Pasadena CA 91101, USA}
% \affiliation{\osuaffil}
% \affiliation{\ccappaffil}
\author[0000-0002-2854-5796]{John D.\ Roberts}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
% \author[0000-0001-7775-7261]{David H. Weinberg}
\author{coauthors}

@@

\begin{abstract}
    Stars in the Milky Way disk exhibit a clear separation into two chemically distinct populations based on [$\alpha$/Fe] ratios. This $\alpha$-bimodality is not a universal feature of simulated disk galaxies and may point to the Milky Way's unique evolutionary history. A popular explanation is the two-infall scenario, which postulates that two periods of substantial accretion rates dominate the assembly history of the Galaxy. 
    However, most previous studies using the two-infall scenario have explored a limited portion of the parameter space, typically neglecting radial migration and assuming that the Galactic disk never ejected a substantial outflow. 
    Thanks to advances in stellar age measurements in recent years, we can now also compare this popular model to more direct measurements of the Galaxy's evolutionary timescales across the disk from large stellar catalogs. 
    We run multi-zone galactic chemical evolution (GCE) models with a two-infall-driven star formation history, radially dependent mass-loaded outflows, and a prescription for radial migration tuned to a hydrodynamical simulation. 
    We compare our model results to abundance patterns across the disk from APOGEE DR17,
    % and early data from the SDSS-V Milky Way Mapper program,
    supplemented with stellar age estimates through multiple methods. 
    Although the two-infall scenario offers a natural explanation for the [$\alpha$/Fe] bimodality, it struggles to explain several features of the age--abundance structure in the disk. The two-infall scenario generically predicts a massive and long-lasting dilution event, but the data show that stellar metallicity is remarkably constant with age across much of the Galactic disk. This apparent age-independence places considerable restrictions upon the two-infall parameter space. These issues can be mitigated, but not completely resolved, by allowing the accreted gas to be pre-enriched to low metallicity. Additionally, the two-infall scenario predicts that local metal-rich stars should have a bimodal distribution of ages, whereas APOGEE data show most of these stars have intermediate ages. Restrictions upon the two-infall parameter space also limit the application of other merger-dominated star formation histories to the Milky Way.
\end{abstract}

@@

% \todo{Outflows.} 
Most previous studies of the two-infall model have not included mass-loaded outflows. Some hydrodynamic simulations of Galactic fountains ejected by CC SNe have shown that ejected material falls back onto the disk on relatively short timescales \citep{spitoni_galactic_2008,spitoni_effects_2009} and close to their point of origin \citep{melioli_hydrodynamical_2008,melioli_hydrodynamical_2009}, suggesting the effect on GCE could be minimal. However, other simulations of MW-like galaxies with different feedback prescriptions do produce mass-loaded outflows \todo{(citations)}. Empirically, mass-loaded outflows have been observed in nearby starburst galaxies \todo{(citations)} and dwarf galaxies \todo{(citations)} but not MW-like systems, although the predicted column densities are below current detection limits \todo{(citations)}. In the absence of outflows, some two-infall studies have implemented radial gas flows or a variable star formation efficiency in order to regulate the radial metallicity gradient \citep{spitoni_effects_2011,palla_chemical_2020}.
% We adopt an {\it ad hoc} prescription for outflows as a function of radius in order to balance different yield sets and reproduce the radial metallicity gradient. \todo{Trim down, move some to methods.}

@@

\begin{table*}
    \centering
    \caption{Median and dispersion in APOGEE parameter uncertainties.}
    \label{tab:uncertainties}
    \begin{tabular}{lcc}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Median Uncertainty} & \multicolumn{1}{c}{Uncertainty Dispersion ($95\%-5\%$)} \\
        \hline
        [Fe/H]          & $0.0089$   & $0.0060$ \\
        $\mathOFe$  & $0.019$    & $0.031$ \\
        $\log_{10}(\tau_{\rm NN}/{\rm Gyr})$    & 0.10   & $0.16$ \\
        $\tau_{\rm [C/N]}/{\rm Gyr}$     & 1.0   & \todo{N/A} \\
        \hline
    \end{tabular}
\end{table*}

@@

@@

Our second age catalog uses the [C/N]--age relation for red giant branch and red clump stars. 
\todo{Paragraph describing [C/N]-derived ages (get help from Jack).}
This method has the benefit of providing age estimates for luminous giants, which increases the sample size at larger distances from the Sun. However, systematic effects of the [C/N]--age relation mean that these ages are most trustworthy in the range $1\sim10$ Gyr. We estimate ages for \num{113464} stars across the disk, including \num{20995} in the Solar neighborhood.
\todo{Note about metallicity cutoff for luminous giants. Cite: \citet{cao_modeling_2025}}

@@

@@

We have

@@

>>>>

% Author list
% \author[]{Amanda Ash}
\author[0000-0003-3781-0747]{Liam O.\ Dubay}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
\author[0000-0001-7258-1834]{Jennifer A.\ Johnson}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
\author[0000-0002-6534-8783]{James W.\ Johnson}
\affiliation{Observatories of the Carnegie Institution for Science, 813 Santa Barbara St., Pasadena CA 91101, USA}
% \affiliation{\osuaffil}
% \affiliation{\ccappaffil}
\author[0000-0002-2854-5796]{John D.\ Roberts}
\affiliation{\osuaffil}
\affiliation{\ccappaffil}
% \author[0000-0001-7775-7261]{David H. Weinberg}
% \author{coauthors}

@@

\begin{abstract}
    Stars in the Milky Way disk exhibit a clear separation into two chemically distinct populations by their [$\alpha$/Fe] ratios. This $\alpha$-bimodality is not a universal feature of simulated disk galaxies and may point to the Milky Way's unique evolutionary history. A popular explanation is the two-infall scenario, which postulates that two periods of substantial accretion rates dominate the assembly history of the Galaxy. 
    However, most previous studies using the two-infall scenario have explored a limited portion of the parameter space, typically neglecting radial migration and assuming that the Galactic disk never ejected a substantial outflow. 
    Thanks to advances in stellar age measurements in recent years, we can now also compare this popular model to more direct measurements of the Galaxy's evolutionary timescales across the disk from large stellar catalogs. 
    We run multi-zone galactic chemical evolution (GCE) models with a two-infall-driven star formation history, radially dependent mass-loaded outflows, and a prescription for radial migration tuned to a hydrodynamical simulation. 
    We compare our model results to abundance patterns across the disk from APOGEE DR17,
    % and early data from the SDSS-V Milky Way Mapper program,
    supplemented with stellar age estimates through multiple methods. 
    Although the two-infall scenario offers a natural explanation for the [$\alpha$/Fe] bimodality, it struggles to explain several features of the age--abundance structure in the disk. The two-infall scenario generically predicts a massive and long-lasting dilution event, but the data show that stellar metallicity is remarkably constant with age across much of the Galactic disk. This apparent age-independence places considerable restrictions upon the two-infall parameter space. These issues can be mitigated, but not completely resolved, by allowing the accreted gas to be pre-enriched to low metallicity. 
    Additionally, the two-infall scenario predicts that local metal-rich stars should have a bimodal distribution of ages, whereas APOGEE data show most of these stars have intermediate ages. Restrictions upon the two-infall parameter space also limit the application of other merger-dominated star formation histories to the Milky Way.
\end{abstract}

@@

% \todo{Outflows.} 
Most previous studies of the two-infall model have not included mass-loaded outflows. Some hydrodynamic simulations of Galactic fountains ejected by CC SNe have shown that ejected material falls back onto the disk on relatively short timescales \citep{spitoni_galactic_2008,spitoni_effects_2009} and close to their point of origin \citep{melioli_hydrodynamical_2008,melioli_hydrodynamical_2009}, suggesting the effect on GCE could be minimal. However, the effects of feedback in simulations are sensitive to its implementation \citep[e.g.,][]{li_effects_2020,hu_code_2023}, and other simulations of MW-like galaxies with different feedback prescriptions do produce mass-loaded outflows \citep[e.g.,][]{brook_hierarchical_2011,gutcke_nihao_2017,nelson_first_2019,peschken_angular_2021,kopenhafer_seeking_2023}. Empirically, mass-loaded outflows have been observed in nearby starburst galaxies \citep[e.g.,][]{lopez_temperature_2020,cameron_duvet_2021,lopez_x-ray_2023} and dwarf galaxies \todo{(citations)} but not MW-like systems, although the predicted column densities are below current detection limits \citep[see reviews by][]{veilleux_cool_2020,thompson_theory_2024}. Even if the MW is not currently ejecting a substantial outflow, it is not unreasonable to suppose that it may have during a more active phase in its evolutionary history.

In the absence of outflows, some two-infall studies have implemented radial gas flows or a variable star formation efficiency in order to regulate the radial metallicity gradient \citep{spitoni_effects_2011,palla_chemical_2020}.
% We adopt an {\it ad hoc} prescription for outflows as a function of radius in order to balance different yield sets and reproduce the radial metallicity gradient. \todo{Trim down, move some to methods.}

@@

\begin{table*}
    \centering
    \caption{Median and dispersion in APOGEE parameter uncertainties.}
    \label{tab:uncertainties}
    \begin{tabular}{lcc}
        \hline\hline
        \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Median Uncertainty} & \multicolumn{1}{c}{Uncertainty Dispersion ($95\%-5\%$)} \\
        \hline
        [Fe/H]          & $0.0089$   & $0.0060$ \\
        $\mathOFe$  & $0.019$    & $0.031$ \\
        $\log_{10}(\tau_{\rm NN}/{\rm Gyr})$    & 0.10   & $0.16$ \\
        $\tau_{\rm [C/N]}/{\rm Gyr}$     & 1.0   & 1.3 \\
        \hline
    \end{tabular}
\end{table*}

@@

\subsection{Stellar Age Estimates}

@@

% Our second age catalog uses the [C/N]--age relation for red giant branch and red clump stars. 
% \todo{Paragraph describing [C/N]-derived ages (get help from Jack).}
% This method has the benefit of providing age estimates for luminous giants, which increases the sample size at larger distances from the Sun. However, systematic effects of the [C/N]--age relation mean that these ages are most trustworthy in the range $1\sim10$ Gyr. We estimate ages for \num{113464} stars across the disk, including \num{20995} in the Solar neighborhood.
% \todo{Note about metallicity cutoff for luminous giants. Cite: \citet{cao_modeling_2025}}

Our second age catalog utilizes the [C/N]--age relation calibrated by \todo{Roberts et al.\ (hopefully submitted)} for red giant branch (RGB) and red clump stars. The relationship relies on the mass-dependent level of mixing during the first dredge-up \citep[FDU;][]{iben_stellar_1967} to map the correlation of stellar mass, and hence age, with surface chemistry. This method has the benefit of providing age estimates for luminous giants, which increases the sample size at larger distances from the Sun. However, limitations from the efficiency of FDU mixing and the RGB age--mass relationship mean the ages are not trustworthy outside the range $1\sim10\Gyr$. Additional mixing effects in low-metallicity stars %later in their RGB evolution 
also prevent the relation from being applied to luminous giant ($\log(g)<2.5$) and red clump stars with $\mathFeH<-0.4$. The median propagated uncertainty for the [C/N]-derived ages is $\sim1\Gyr$; however, as noted by \todo{Roberts et al.\ (hopefully submitted)}, the propagated errors underestimate the true age dispersion and so we enhance these uncertainties by 40\% to account for this \todo{(rewrite)}. With this relationship, we estimate ages for \num{113464} stars across the disk, including \num{20995} in the Solar neighborhood.

@@

\subsection{Comparison with Previous Literature}

@@

We have compared the predictions of the two-infall scenario against abundance data from APOGEE DR17 supplemented with age estimates using two different methods. We ran multi-zone GCE models at two different yield scales with prescriptions for radially-dependent outflows and stellar migration. While the two-infall scenario can explain the local stellar abundance distribution, in particular the $\alpha$-bimodality, it faces challenges in matching the age--abundance structure of the full disk. We explored multiple parameter modifications to bring the model predictions closer to the data, including the yield scale, radial migration strength, metallicity of the accreted gas, thick-to-thin disk mass ratio, and the SN Ia DTD. Our conclusions are as follows:

@@

The apparent age-independence of stellar abundances in the disk places considerable restrictions upon the two-infall parameter space because it predicts a substantial dilution event at the start of the thin disk epoch. If the equilibrium scenario of \citet{johnson_milky_2024} is correct, then it restricts the two-infall scenario more than other GCE models.

\todo{Implications for merger-driven SFHs.}
